## Data visualization

By joining attributes with spatial boundaries, we can map demographic patterns. When it comes to mapping, it is important to consider what type of variable you intend to display. In this tutorial, we will cover two different types of maps: *Choropleth map* for quantitative variables and *Typology map* for qualitative variables.

### Choropleth map

A choropleth map is useful for visualizing quantitative variables. This type of map displays color corresponding with an aggregated variable for a given geographic unit, such as population density or per-capital income. Below, this tutorial will walk you through how to create a demographic variable of interest, and visualize it using a choropleth map.

In social science research, scholars often use *the Herfindahl-Hirschman Index (HHI)* as a measure for *diversity*. It is an economic index measuring the level of market competition/monopolization, but the formula is also used in ecology (called the Simpson diversity index), as well as in demographic contexts. The formula is:

$$
HHI = \sum_{i=1}^{N} (MS_i)^2
$$
Where $MS_i$ is the market share of firm $i$ in the market, and $N$ is the number of firms. The lower the HHI, the more competition in the market (many firms with relatively equal market shares). Reversely, the higher the HHI, the less competition in the market, suggesting monopoly.

In demographic research, sociologists commonly use the HHI to measure racial diversity of neighborhoods. Instead of market share of firms in a market, we use the shares of racial groups in a neighborhood.

For example, in a neighborhood with 4 racial groups each representing 25%, the HHI would be:

$$
0.25^2 + 0.25^2 + 0.25^2 + 0.25^2 = 0.25
$$

Compared this to another neighborhood with 2 racial groups -- one representing 80% and the other representing 20%. The HHI would be:

$$
0.80^2 + 0.20^2 = 0.68 
$$

The first neighborhood with the HHI of 25% is more diverse compared to the second neighborhood with the HHI of 68%.

We can create this diversity measure using the variables we have.

```{r make-hhi}
# Create the HHI 
sf_joined <- sf_joined %>% 
  mutate(diversity = (pwhite^2 + pblack^2 + phisp^2 + pasian^2)/100)

# Check
sf_joined %>%
  select(diversity)
```

Next, we can create a map displaying how diverse/homogeneous San Francisco neighborhoods are. 

```{r choro-map, warning=FALSE, message=FALSE}
# Create a choropleth map visualizing the diversity variable
ggplot(data = sf_joined) + # specify the source of spatial data
  geom_sf(aes(fill = diversity), 
          size = 0.2,
          color = "white") + # add a layer to use "diversity" to fill colors
  scale_fill_distiller(type="seq", # fill in sequential colors 
                       palette = "Greys",
                       direction = -1, # 1 for low (light) to high (dark), -1 for reverse
                       ) +
  geom_sf_label(aes(label = nhood),  # add labels using the nhood variable
                 size = 1.5, # size of the label text
                 color = "black",         # color of the text
                 fill = "white",          # background fill color
                 label.padding = unit(0.1, "lines"),  # padding around the text
                 label.size = 0,        # thickness of the border around the label
                 alpha = 0.5,             # set transparency of the background
                 check_overlap = TRUE) + # suppress overlaping texts
  theme_void() +
  labs(fill = "HHI", # add a label for legend
       title = "Gentrification in San Francisco" # figure title
       ) 
```


 - What are the least diverse neighborhoods? Which is the dominant racial group? 
 - What are the most diverse neighborhoods?
 

### Typology map

A typology map is suitable for qualitative variables that consist distinct and meaningful categories. We will create a gentrification measure to demonstrate a use case for typology map. Gentrification is a phenomenon that we measure over space and time. While there are multiple ways to define and measure gentrification, for the purposes of this tutorial, we will conceptualize gentrification as a neighborhood-level socioeconomic transformation comprised of both an influx of middle-and upper-class residents and an increase in housing prices in previously low-income, urban neighborhoods.

Our measure of gentrification is calculated at the census tract level from 2000 to 2020, so that we can examine how the socioeconomic composition of neighborhoods in San Francisco changes over time. The four variables listed below are used in our categorical measure of gentrification: 

 * Median household income (in 2020 Dollars)
 * Percent of college-educated residents 
 * Median home value (in 2020 Dollars)
 * Median gross rent (in 2020 Dollars)  
  
The first step in computing the gentrification measure is determining whether or not a tract is **eligible to gentrify (i.e., Gentrifiable).** Generally, researchers consider tracts eligible to gentrify if they are relatively low-income neighborhoods such that they could undergo the revitalization that characterizes gentrification. We will operationalize eligibility to gentrify in a binary fashion using the median household income.

>  **Not Gentrifiable**: Census tracts had a median household income *above* the city-wide median household income at the start of the period. 

> **Gentrifiable**: Census tracts had a median household income *below* the city-wide median household income at the start of the period. 

Next, among the census tracts that were Gentrifiable, we must determine if they were gentrifying over time or not. To classify the gentrifiable tracts as gentrifying or not gentrifying, we will need to calculate how much socioeconomic change and housing value change the tract experienced over the decade and compare the rate of change to the rate in the city. 

> **Gentrifying**: Gentrifiable census tracts that had the socioeconomic status of their residents (i.e., % college graduates OR median household income) grow faster than the city wide-median *AND* the home values (median home value OR median gross rent value) grow faster than the city-wide median over the decade period. 

> **Not Gentrifying**: Gentrifiable census tracts that did NOT experience growth in the socioeconomic status of their residents *AND* their home values at a rate higher than the city-wide median values over the decade period.

This measurement strategy results in a three-category gentrification measure - Not Gentrifiable, Gentrifying, and Not Gentrifying. With these three categories we can make a typology map of census tracts in San Francisco.

```{r import-trt}
# Import demographic attributes for census tracts
sftrt_dem <- read.csv("data/sftrt_dem.csv")

# Check if it has a common identifier with sftrt
head(sftrt_dem$trt10)
class(sftrt_dem$trt10)

head(sftrt$tractce10)
class(sftrt$tractce10)

# Fill with leading zeros and convert to string
sftrt_dem <- sftrt_dem %>%
  mutate(tractce10 = as.character(sprintf("%06d", trt10)))

# Join demographic data with geometric boundaries
sftrt_joined <- merge(
  x = sftrt,               # sf object
  y = sftrt_dem,           # data frame
  by.x = "tractce10",      # x identifier
  by.y = "tractce10",      # y identifier
  all.x = TRUE,            # keep all lines
)

```

```{r typo-map}
# Create a typology map
ggplot(data = sftrt_joined) + 
  geom_sf(aes(fill = gentcat), 
          size = 0.02,
          color = "white") +
  scale_fill_manual(values = c("non-gentrifiable" = "light grey", 
                               "gentrifying" = "pink", 
                               "non-gentrifying" = "black")) +
  theme_void() + # minimalist theme
  labs(fill = "Gentrification Category", # legend label
       title = "Gentrification in San Francisco" # figure title
       ) + 
  # customize legend position and map layout (optional)
  theme(legend.position = "right")  # You can also use "bottom" or "left")
```


