# From People to Built Environment


So far, we have focused on embedding demographic data into places. However, sociological questions extend beyond just understanding people in places. People don’t simply exist in locations—they interact with the built environment, including schools, local businesses, landmarks, and more. In GIS terms, this type of data is known as point-of-interest (POI) data, which refers to specific points or useful sites identified by geographic coordinates (latitude and longitude). In this tutorial, we will use Airbnb listings and local business data as examples of point-of-interest data.

## Transforming Airbnbs

When data contains latitude and longitude coordinates, we can transform it into an sf object by using the st_as_sf command. Here, we’ll use San Francisco Airbnb data to illustrate this process. Unlike neighborhood boundaries, which have polygon geometries, POIs will have **point** geometries.

```{r import-bnb}
# Import the Airbnb data
sfbnb <- read.csv("data/sfbnb.csv")

head(sfbnb, 3)
```

```{r transform-bnb}
# Transform lat/long into sf object
sf_sfbnb <- st_as_sf(sfbnb, 
                     coords = c("longitude", "latitude"), 
                     crs = 4326 # specify the projection; WGS84 is a standard projection for global mapping. 
                     )
# Check the data
head(sf_sfbnb, 3)
```

You can see that geometry is points, rather than polygons. Each point represents an Airbnb property (id- unique identifier). It also indicates the census tract (trt10) and neighborhood name (analysis) in which the Airbnb  property is located. The data also contains attributes of the Airbnb property, such as property type, review score ratings, and the number of reviews.

You can save the converted file as geojson using the "st_write" command.

```{r export-bnb, results='hide', warning=FALSE, message=FALSE}
# Export the Airbnb data as a geojson file
st_write(sf_sfbnb, "processed-data/airbnbs.geojson", driver = "GeoJSON", delete_dsn = TRUE)
```


## Geocoding Restaurants

Sometimes, your data may lack latitude and longitude information. Instead, the data may contain street addresses. In such cases, we need to perform **geocoding** which converts formatted addresses into latitude and longitude coordinates, allowing the data to be displayed as points on a map. Below, we will demonstrate the geocoding process using the restaurant data in San Francisco.

```{r import-biz}
# Import restaurant data
sfbiz <- read.csv("data/sfbiz_clean.csv") 

# Check if lat and lon exists
head(sfbiz, 3)
```

In this data, we can see that latitude and longitude coordinates do not exist. However, it contains information on street address, city, state, and zip code. To carry out geocoding, first, we have to create a field that displays a full address. Based on the full address, we can get the coordinates via open street map (osm). Once the data is geocoded, we can transform this data frame to sf object for mapping.

```{r geocode, warning=FALSE, message=FALSE}
# Set up
library(tidygeocoder)

# Create a full address field (street address, city, state, zipcode)
sfbiz <- sfbiz %>%
  mutate(full_add = paste(address_line_1, city, state, zipcode, sep = ", "))

# Geocode the full address
geocoded_sfbiz <- sfbiz %>%
  sample_n(5) %>% # the full data can take a while, so let's try on a smaller sample
  geocode(address = full_add, method = 'osm')

# Check the geocoded data
geocoded_sfbiz %>%
  dplyr::select(lat, long)
```

```{r import-geocoded}
# For the purpose of this tutorial, we can use the prepared, full geocoded data
prep_sfbiz <- read.csv("data/geocoded_sfbiz_clean.csv") 

# Transform lat/long data into sf oject
sf_sfbiz <- st_as_sf(prep_sfbiz, 
                     coords = c("long", "lat"), 
                     crs = 4326)

head(sf_sfbiz, 5)
```

```{r export-biz, results='hide', warning=FALSE, message=FALSE}
# Export the restaurant data as a geojson file
st_write(sf_sfbiz, "processed-data/restaurants.geojson", driver = "GeoJSON", delete_dsn = TRUE)
```


## Distribution and density

Now that we have restaurant locations as points, we can simply overlay the restaurant locations on top of the neighborhood boundaries layer to explore the geographical distribution of restaurants.

```{r point-map}
# Plot the restaurant locations
ggplot() +
  # display neighborhood boundaries as a layer
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  # add restaurants as another layer
  geom_sf(data = sf_sfbiz, 
          color = "red",    
          size = 0.5,       
          alpha = 0.8       # set transparency
          ) +  
  theme_void() +
  labs(title = "Restaurants in San Francisco")
```


**SYMBOLOGY BASED ON ATTRIBUTES**

You can differentiate points based on attributes using symbology (e.g., colors, shapes, sizes). Here, we are only looking at restaurants. But imagine you have a dataset that contains multiple types of retail, such as bars, coffee shops, and restaurants. You might want to map the overall distribution of businesses, but also focus on one specific type of business or highlight differences by business type.

In our data, we have "employee size" and "sales volume" of restaurants as attributes. Let's say you want to focus on small, mom and pop businesses and explore where these small restaurants are concentrated. In the US, businesses with less than 10 or 5 employees are often considered as "small". Here, we will use 5 as a cutoff to define a small restaurant. You can highlight your data based on this specific attribute.

```{r point-attribute-map}
# Create a new column based on employee size
sf_sfbiz$small <- ifelse(sf_sfbiz$employee_size_location < 5, "small", "non-small")

# Plot with ggplot
ggplot() +
  # display neighborhood boundaries as a layer
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  # add restaurants as another layer
  geom_sf(data = sf_sfbiz,
          aes(color = small), # specify the color based on "small" variable
          size = 0.5, alpha = 0.8
          ) +
  scale_color_manual(values = c("small" = "red", "non-small" = "black"), # specify colors 
                     labels = c("small" = "less than 5", 
                                "non-small" = "5 or more")
                                ) +           # set labels
  theme_void() +
  labs(title = "Restaurants by Employee Size in San Francisco",
       color = "Employee Size")
```

```{r add-symbology}
# Add more symbology details
ggplot() +
  # display neighborhood boundaries as a layer
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  # add restaurants as another layer
  geom_sf(data = sf_sfbiz,
          aes(color = small,
              shape = small,
              size = small), 
          alpha = 0.8) +
  scale_color_manual(values = c("small" = "red", "non-small" = "black")) +  
  scale_shape_manual(values = c("small" = 23, "non-small" = 16),
                     labels = c("small" = "less than 5", 
                                "non-small" = "5 or more")) + 
  scale_size_manual(values = c("small" = 1.5, "non-small" = 0.5)) + 
  guides(color = "none",
         size = "none",
         shape = guide_legend(override.aes = list(
           color = c("red", "black"), 
           size = c(1.5, 0.5), 
           shape = c(23, 16)
           ))) +
  theme_void() +
  labs(title = "Restaurants by Employee Size in San Francisco",
       shape = "Employee Size")
```


From point maps, we can get a general sense of the distribution of restaurants in San Francisco. However, there are two limitations to this visualization. First, it is hard to tell exactly where restaurants are more or less concentrated because points overlap where densely populated. To draw attention to where points are concentrated the most, we can use a **density map** (also known as a heat map). Among various methods for creating a density map, we will learn how to create a density map using kernel density estimation (KDE). This function is available through the "MASS" package.


```{r density-map, message=FALSE}
# Install and set up the "MASS" package 
library("MASS")

# Extract coordinates from the sf object
coords <- st_coordinates(sf_sfbiz)

# Calculate KDE based on coordinates
kde <- kde2d(coords[, 1], coords[, 2], n = 500) # n specifies the number of grids

# Convert KDE to raster format
kde_df <- expand.grid(x = kde$x, y = kde$y)
kde_df$z <- as.vector(kde$z)

head(kde_df, 5)

# Plot KDE heat map
ggplot() +
  geom_raster(data = kde_df, 
              aes(x = x, y = y, fill = z), 
              alpha = 1) +  # Kernel density heat map
  geom_sf(data = sfnh, fill = NA, color = "black") +  # Neighborhood boundaries
  scale_fill_gradientn(colors = c("transparent", "lightpink", "red", "darkred"), name = "Density") +
  theme_void() +
  labs(title = "Heat Map of Restaurants in San Francisco")
```

Now, you can clearly see where the restaurants are most concentrated!

While heat maps are useful for clearly visualizing concentration of restaurants, the second limitation still remains. With points data, it is difficult to discern spatial patterns of restaurants by neighborhoods because restaurants are both in and outside of neighborhood boundaries. To solve this problem, we will discuss **spatial join** in the following section, which allows us to spatially link restaurants to neighborhoods, so that we can reduce points data to the level of neighborhoods.



**CODING EXERCISE**

 - Create a point map using symbology
   - Use the sales volume variable to create a new category
   - Try using different colors, shapes, and sizes to represent your data


