## Distribution and density

```{r import-processed-map}
sfnh <- st_read("data/sfnh.geojson")
sf_sfbiz <- st_read("processed-data/restaurants.geojson")
```

Now that we have restaurant locations as points, we can simply overlay the restaurant locations on top of the neighborhood boundaries layer to explore the geographical distribution of restaurants.

```{r point-map}
# Plot the restaurant locations
ggplot() +
  # display neighborhood boundaries as a layer
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  # add restaurants as another layer
  geom_sf(data = sf_sfbiz, 
          color = "red",    
          size = 0.5,       
          alpha = 0.8       # set transparency
          ) +  
  theme_void() +
  labs(title = "Restaurants in San Francisco")
```


**SYMBOLOGY BASED ON ATTRIBUTES**

You can differentiate points based on attributes using symbology based on colors or shapes. Here, we are only looking at restaurants. But imagine you have a dataset that contains multiple types of retail, such as bars, coffee shops, and restaurants. You might want to map the overall distribution of businesses, but also focus on one specific type of business or highlight differences by business type.

In our data, we have "employee size" and "sales volume" of restaurants as attributes. Let's say you want to focus on small, mom and pop businesses and explore where these small restaurants are concentrated. In the US, businesses with less than 10 or 5 employees are often considered as "small". Here, we will use 5 as a cutoff to define a small restaurant. You can highlight your data based on this specific attribute.

```{r}
# Create a new column based on employee size
sf_sfbiz$small <- ifelse(sf_sfbiz$employee_size_location < 5, "small", "non-small")

# Plot with ggplot
ggplot() +
  # display neighborhood boundaries as a layer
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  # add restaurants as another layer
  geom_sf(data = sf_sfbiz,
          aes(color = small), # specify the color based on "small" variable
          size = 0.5, alpha = 0.8
          ) +
  scale_color_manual(values = c("small" = "red", "non-small" = "black"), # specify colors 
                     labels = c("less than 5", "5 or more")) +           # set labels
  theme_void() +
  labs(title = "Restaurants by Employee Size in San Francisco",
       color = "Employee Size")
```

```{r}
# Add more symbology details
ggplot() +
  # display neighborhood boundaries as a layer
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  # add restaurants as another layer
  geom_sf(data = sf_sfbiz,
          aes(color = small,
              shape = small,
              size = small), # specify the color based on "small" variable
          alpha = 0.8) +
  scale_color_manual(values = c("small" = "red", "non-small" = "black")) +  
  scale_shape_manual(values = c("small" = 23, "non-small" = 16),
                     labels = c("less than 5", "5 or more")) + 
  scale_size_manual(values = c("small" = 1.5, "non-small" = 0.5)) + 
  guides(color = "none",
         size = "none",
         shape = guide_legend(override.aes = list(
           color = c("red", "black"), 
           size = c(1.5, 0.5), 
           shape = c(23, 16)
           ))) +
  theme_void() +
  labs(title = "Restaurants by Employee Size in San Francisco",
       shape = "Employee Size")
```

In this map, we can get a general sense of the distribution of restaurants in San Francisco. However, there are two limitations to this visualization. First, it is hard to tell exactly where restaurants are more or less concentrated because points overlap where densely populated. To draw attention to where points are concentrated the most, we can use a **density map** (also known as a heat map). Among various methods for creating a density map, we will learn how to create a density map using kernel density estimation (KDE). This function is available through the "MASS" package.


```{r density-map}
# Install and set up the "MASS" package 
library("MASS")

# Extract coordinates from the sf object
coords <- st_coordinates(sf_sfbiz)

# Calculate KDE based on coordinates
kde <- kde2d(coords[, 1], coords[, 2], n = 500) # n specifies the number of grids

# Convert KDE to raster format
kde_df <- expand.grid(x = kde$x, y = kde$y)
kde_df$z <- as.vector(kde$z)

# Plot KDE heat map
ggplot() +
  geom_raster(data = kde_df, 
              aes(x = x, y = y, fill = z), 
              alpha = 1) +  # Kernel density heat map
  geom_sf(data = sfnh, fill = NA, color = "black") +  # Neighborhood boundaries
  scale_fill_gradientn(colors = c("transparent", "lightpink", "red", "darkred"), name = "Density") +
  theme_void() +
  labs(title = "Heat Map of Restaurants in San Francisco")

```

Now, you can clearly see where the restaurants are most concentrated!

While heat maps are useful for clearly visualizing concentration of restaurants, the second limitation still remains. With points data, it is difficult to discern spatial patterns of restaurants by neighborhoods because restaurants are both in and outside of neighborhood boundaries. To solve this problem, we will discuss **spatial join** in the following section, which allows us to spatially link restaurants to neighborhoods, so that we can reduce points data to the level of neighborhoods.


## Spatial join

Spatial join allows you to combine two sf objects based on the spatial relationship between their geometries. For example, we can think of the relationship between neighborhoods (polygons) and restaurants (points).

 * Neighborhoods (x) contain restaurants (y), or
 * restaurants (x) are within neighborhoods (y).

```{r spatial-join}
# Before joining, check if they have the same projections
st_crs(sfnh) == st_crs(sf_sfbiz)

# Perform spatial join
nh_joined <- st_join(x = sfnh, # join
                     y = sf_sfbiz, # target
                     join = st_contains, # does x(polygon) contains y(point)?
                     left = TRUE) # keep all neighborhoods

# Explore
head(nh_joined, 3)
```

In the joined data, we can see that restaurants (company) are nested within neighborhoods (nhood). We can aggregate this data to the neighborhood level and get the number of restaurants per neighborhood. Some neighborhoods may not any have restaurants, so you want to preserve that as 0 and not count as 1.

```{r count-biz}
# Count restaurants per neighborhood
restaurant_counts <- nh_joined %>%
  st_drop_geometry() %>%
  group_by(nhood) %>% 
  summarise(n_rst = sum(!is.na(company))) # don't count NA as 1

head(restaurant_counts, 10)
```

**MAPPING AGGREGATED RESTAURANT COUNTS**

As the count of restaurants is a quantitative variable, we can visualize it using the choropleth map. I first join the restaurant counts data with neighborhood boundaries to get the neighborhood-level spatial data. Then, I create a map using custom breaks based on the distribution of the number of restaurants within a neighborhood. 

```{r aggregate-join}
# Join the restaurant count to neighborhood boundaries
biz_colors <- sfnh %>%
  left_join(restaurant_counts,
            by = "nhood")

head(biz_colors, 3)
hist(biz_colors$n_rst)
```

```{r aggregate-map}
# Create the map using Custom breaks
ggplot() +
  geom_sf(data = biz_colors,
          aes(fill = n_rst), 
          size = 0.2,
          color = "white") +  
  scale_fill_distiller(type="seq", 
                       palette = "Greys",
                       breaks = c(50, 150, 250), # specify custom breaks
                       direction = 1,
                       ) +
  theme_void() +
  labs(fill = "Restaurant Count",
       title = "Choropleth Map of Restaurants in San Francisco")
```

Another way of visualizing a quantitative variable is a *proportional symbol map*. Proportional symbol maps are particularly useful for absolute quantitative variables (e.g., raw counts, measurements) while choropleth maps are more suitable for relative quantitative variables (e.g., ratios, percentages, proportions).

```{r proportional-symbol-map, warning=FALSE}
# Create neighborhood centroids
nh_cent <- st_centroid(sfnh)

# Check centroids
print(nh_cent)

# Join the restaurant counts to neighborhood centroids
biz_symbols <- nh_cent %>%
  left_join(restaurant_counts, by = "nhood") %>%
  arrange(desc(n_rst)) # sort to ensure small points would be plotted in front of big points

head(biz_symbols, 3)

# Create the proportional symbol map
ggplot() +
  geom_sf(data = sfnh,           # add a base map layer of boundaries
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = biz_symbols,
          aes(size = n_rst),      # add a layer of symbols sized based on the restaurant counts 
          shape = 21,             # specify a circle shape for symbols
          fill = "red",           # set a color to fill the shape 
          alpha = 0.6,            # set a level of transparency
          color = "lightgray") +  # set a color for edges of the shape
  scale_size(range = c(3, 13)) +  # set min and max for the size of the symbols
  theme_void() +
  labs(size = "Restaurant Count",
       title = "Proportional Symbol Map of Restaurants in San Francisco") 
```

A proportional symbol map is also called a clustered point map. In this map, the symbol sizes correspond with the frequency of restaurants - the larger the symbol, the more restaurants that are present in that area. This type of clustering and aggregating visualization technique is particularly helpful when your data only have approximate locations. You can combine two types of aggregated into one visualization by specifying both symbol sizes and colors.

```{r}
# Create the proportional symbol map (size) combined with the choropleth map (fill color)
ggplot() +
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = biz_symbols,
          aes(size = n_rst, fill = n_rst), 
          shape = 21, 
          color = "lightgray") +
  scale_size(range = c(3, 10)) +
  # add colors to the symbols
  scale_fill_gradientn(colors = hcl.colors(4,                   # fill the shape with four different colors
                                           "RdBu",              # red-blue color scheme
                                           rev = TRUE,          # reverse the color scheme
                                           alpha = 0.9)) +      # transparency
  theme_void() +
  guides(fill = guide_legend(title = "Restaurant Count"), 
         size = guide_legend(title = "Restaurant Count")) +     # combine legends
  labs(title = "Choropleth/Proportional Symbol Map of Restaurants in San Francisco")
```


## Export Maps

Now, let's say you are ready export your maps for publication. We will publish a figure of two maps side by side, displaying two ways of visualizing the distribution of restaurant by neighborhood in San Francisco.

```{r store-and-export-maps}
# Store and combine maps
map1 <- ggplot() +
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = sf_sfbiz, 
          color = "red", 
          size = 0.5, alpha = 0.8
          ) + 
  theme_void()

map2 <- ggplot() +
  geom_raster(data = kde_df, 
              aes(x = x, y = y, fill = z), 
              alpha = 1) +  
  geom_sf(data = sfnh, fill = NA, color = "black") +  
  scale_fill_gradientn(colors = c("transparent", "lightpink", "red", "darkred"), name = "Density") +
  theme_void() +
  theme(legend.position = "none")

map3 <- ggplot() +
  geom_sf(data = biz_colors,
          aes(fill = n_rst), 
          size = 0.2,
          color = "white") +  
  scale_fill_distiller(type="seq", 
                       palette = "Greys",
                       breaks = c(50, 150, 250), 
                       direction = 1,
                       ) +
  theme_void() +
  theme(legend.position = "none")

map4 <- ggplot() +
  geom_sf(data = sfnh,          
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = biz_symbols,
          aes(size = n_rst),      
          shape = 21,           
          fill = "red",           
          alpha = 0.6,           
          color = "lightgray") + 
  scale_size(range = c(3, 13)) +  
  theme_void() +
  theme(legend.position = "none")

combined <- ggpubr::ggarrange(map1, map2, map3, map4, nrow=2, ncol=2,
                  labels = c("Point Map", "Heat Map", "Choropleth Map", "PropSymbol Map"))

print(combined)

# Export the combined map
ggsave("sf_restaurant_maps.png", plot = combined, 
       width = 9, height = 9, 
       dpi = 300 # resolution
       )
```

**THINK AND SHARE**

Discuss similarities and differences across four different types of maps. What are pros and cons of each type of map? Which visualization do you think is most effective and why?


While it is not strictly expected in sociology papers, you can also add a north arrow and a scale bar to a map using the ggspatial package.

```{r annotate-map, warning=FALSE, message=FALSE}
library(ggspatial)
```

```{r add-arrow-scale}
# Adding a scale bar and a north arrow 
ggplot() +
  geom_sf(data = biz_colors,
          aes(fill = n_rst), 
          size = 0.2,
          color = "white") +  
  scale_fill_distiller(type = "seq", 
                       palette = "Greys",
                       breaks = c(50, 150, 250),
                       direction = 1) +
  theme_void() +
  labs(fill = "Restaurant Count",
       title = "Distribution of Restaurants in San Francisco") +
  # add a scale bar
  annotation_scale(location = "bl",  # "br" is for bottom right, adjust as needed
                   pad_y = unit(0.01, "cm") # place the scale bar close to the bottom
                   ) + 
  # add a north arrow
  annotation_north_arrow(location = "tl",  # "tl" is for top left, adjust as needed
                         style = north_arrow_fancy_orienteering)  
```


**LEARN MORE**

For visualizing density, **hotspot maps** also exist though we do not cover in this tutorial. What makes a hotspot map different from a heat map? The short answer is statistics! A heat map provides a smooth, continuous visualization of data across the given area of interest (e.g., neighborhood), similar to histogram. A hotspot map highlights areas that are statistically significant by comparing the distribution of points in the area of interest (e.g., neighborhood) to how they would be distributed in a random manner. For example, a hotspot map of crimes would demonstrate which areas have more or less incidents than we would expect if the “crime generating process” was completely random. The hotspots indicate areas where the frequency of crimes is higher than we would expect if crime was randomly distributed across the neighborhood. If you want to learn how to make a hotspot map, see [this tutorial](https://crimebythenumbers.com/hotspot-maps.html#making-a-hotspot-map).


