## Spatial join

Spatial join allows you to combine two sf objects based on the spatial relationship between their geometries. For example, we can think of the relationship between neighborhoods (polygons) and restaurants (points).

 * Neighborhoods (x) contain restaurants (y), or
 * restaurants (x) are within neighborhoods (y).

```{r spatial-join}
# Before joining, check if they have the same projections
st_crs(sfnh) == st_crs(sf_sfbiz)

# Perform spatial join
nh_joined <- st_join(x = sfnh, # join
                     y = sf_sfbiz, # target
                     join = st_contains, # does x(polygon) contains y(point)?
                     left = TRUE) # keep all neighborhoods

# Explore
head(nh_joined, 3)
```

In the joined data, we can see that restaurants (company) are nested within neighborhoods (nhood). We can aggregate this data to the neighborhood level and get the number of restaurants per neighborhood. Some neighborhoods may not any have restaurants, so you want to preserve that as 0 and not count as 1.

```{r count-biz}
# Count restaurants per neighborhood
restaurant_counts <- nh_joined %>%
  st_drop_geometry() %>%
  group_by(nhood) %>% 
  summarise(n_rst = sum(!is.na(company))) # don't count NA as 1

head(restaurant_counts, 10)
```


**MAPPING AGGREGATED RESTAURANT COUNTS**

As the count of restaurants is a quantitative variable, we can visualize it using the choropleth map. I first join the restaurant counts data with neighborhood boundaries to get the neighborhood-level spatial data. Then, I create a map using custom breaks based on the distribution of the number of restaurants within a neighborhood. 

```{r aggregate-join}
# Join the restaurant count to neighborhood boundaries
biz_colors <- sfnh %>%
  left_join(restaurant_counts,
            by = "nhood")

head(biz_colors, 3)
hist(biz_colors$n_rst)
```

```{r save aggregated data, message=FALSE}
st_write(biz_colors, "processed-data/agg_restaurants.geojson", driver = "GeoJSON", delete_dsn = TRUE)
```

```{r aggregate-map}
# Create the map using custom breaks
ggplot() +
  geom_sf(data = biz_colors,
          aes(fill = n_rst), 
          size = 0.2,
          color = "white") +  
  scale_fill_distiller(type="seq", 
                       palette = "Greys",
                       breaks = c(50, 150, 250), # specify custom breaks
                       direction = 1,
                       ) +
  theme_void() +
  labs(fill = "Restaurant Count",
       title = "Choropleth Map of Restaurants in San Francisco")
```

Another way of visualizing a quantitative variable is a *proportional symbol map*. Proportional symbol maps are particularly useful for absolute quantitative variables (e.g., raw counts, measurements) while choropleth maps are more suitable for relative quantitative variables (e.g., ratios, percentages, proportions).

```{r proportional-symbol-map, warning=FALSE}
# Create neighborhood centroids
nh_cent <- st_centroid(sfnh)

# Check centroids
print(nh_cent)

# Join the restaurant counts to neighborhood centroids
biz_symbols <- nh_cent %>%
  left_join(restaurant_counts, by = "nhood") %>%
  arrange(desc(n_rst)) # sort to ensure small points would be plotted in front of big points

head(biz_symbols, 3)

# Create the proportional symbol map
ggplot() +
  geom_sf(data = sfnh,           # add a base map layer of boundaries
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = biz_symbols,
          aes(size = n_rst),      # add a layer of symbols sized based on the restaurant counts 
          shape = 21,             # specify a circle shape for symbols
          fill = "red",           # set a color to fill the shape 
          alpha = 0.6,            # set a level of transparency
          color = "lightgray") +  # set a color for edges of the shape
  scale_size(range = c(3, 13)) +  # set min and max for the size of the symbols
  theme_void() +
  labs(size = "Restaurant Count",
       title = "Proportional Symbol Map of Restaurants in San Francisco") 
```

A proportional symbol map is also called a clustered point map. In this map, the symbol sizes correspond with the frequency of restaurants - the larger the symbol, the more restaurants that are present in that area. This type of clustering and aggregating visualization technique is particularly helpful when your data only have approximate locations. You can combine two types of aggregated into one visualization by specifying both symbol sizes and colors.

```{r proportional-choropleth}
# Create the proportional symbol map (size) combined with the choropleth map (fill color)
ggplot() +
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = biz_symbols,
          aes(size = n_rst, fill = n_rst), 
          shape = 21, 
          color = "lightgray") +
  scale_size(range = c(3, 10)) +
  # add colors to the symbols
  scale_fill_gradientn(colors = hcl.colors(4,                   # fill the shape with four different colors
                                           "RdBu",              # red-blue color scheme
                                           rev = TRUE,          # reverse the color scheme
                                           alpha = 0.9)) +      # transparency
  theme_void() +
  guides(fill = guide_legend(title = "Restaurant Count"), 
         size = guide_legend(title = "Restaurant Count")) +     # combine legends
  labs(title = "Choropleth/Proportional Symbol Map of Restaurants in San Francisco")
```


## Export Maps

Now, let's say you are ready export your maps for publication. We will publish a figure of four maps combined, displaying various ways of visualizing the distribution of restaurant by neighborhood in San Francisco.

```{r store-and-export-maps}
# Store and combine maps
map1 <- ggplot() +
  geom_sf(data = sfnh,
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = sf_sfbiz, 
          color = "red", 
          size = 0.5, alpha = 0.8
          ) + 
  theme_void()

map2 <- ggplot() +
  geom_raster(data = kde_df, 
              aes(x = x, y = y, fill = z), 
              alpha = 1) +  
  geom_sf(data = sfnh, fill = NA, color = "black") +  
  scale_fill_gradientn(colors = c("transparent", "lightpink", "red", "darkred"), name = "Density") +
  theme_void() +
  theme(legend.position = "none")

map3 <- ggplot() +
  geom_sf(data = biz_colors,
          aes(fill = n_rst), 
          size = 0.2,
          color = "white") +  
  scale_fill_distiller(type="seq", 
                       palette = "Greys",
                       breaks = c(50, 150, 250), 
                       direction = 1,
                       ) +
  theme_void() +
  theme(legend.position = "none")

map4 <- ggplot() +
  geom_sf(data = sfnh,          
          fill = "lightgray",
          size = 0.02,
          color = "white"
          ) +
  geom_sf(data = biz_symbols,
          aes(size = n_rst),      
          shape = 21,           
          fill = "red",           
          alpha = 0.6,           
          color = "lightgray") + 
  scale_size(range = c(3, 13)) +  
  theme_void() +
  theme(legend.position = "none")

combined <- ggpubr::ggarrange(map1, map2, map3, map4, nrow=2, ncol=2,
                  labels = c("Point Map", "Heat Map", "Choropleth Map", "PropSymbol Map"))

print(combined)

# Export the combined map
ggsave("sf_restaurant_maps.png", plot = combined, 
       width = 9, height = 9, 
       dpi = 300 # resolution
       )
```


**THINK AND SHARE**

Discuss similarities and differences across four different types of maps. What are pros and cons of each type of map? Which visualization do you think is most effective and why?



**LEARN MORE**

While it is not strictly expected in sociology papers, you can also add a north arrow and a scale bar to a map using the ggspatial package.

```{r annotate-map, warning=FALSE, message=FALSE}
library(ggspatial)
```

```{r add-arrow-scale}
# Adding a scale bar and a north arrow 
ggplot() +
  geom_sf(data = biz_colors,
          aes(fill = n_rst), 
          size = 0.2,
          color = "white") +  
  scale_fill_distiller(type = "seq", 
                       palette = "Greys",
                       breaks = c(50, 150, 250),
                       direction = 1) +
  theme_void() +
  labs(fill = "Restaurant Count",
       title = "Distribution of Restaurants in San Francisco") +
  # add a scale bar
  annotation_scale(location = "bl",  # "br" is for bottom right, adjust as needed
                   pad_y = unit(0.01, "cm") # place the scale bar close to the bottom
                   ) + 
  # add a north arrow
  annotation_north_arrow(location = "tl",  # "tl" is for top left, adjust as needed
                         style = north_arrow_fancy_orienteering)  
```


